Medical Imaging Classification with Cold-Atom Reservoir Computing using Auto-Encoders and Surrogate-Driven Training\\

### 2025-05-29 paper/main.tex
Updates from Overleaf

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{float}
\usepackage{inputenc}

@@

% \and
% \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% \textit{name of organization (of Aff.)}\\
% City, Country \\
% email address or ORCID}
% }

@@

\begin{abstract}
We introduce a quantum-classical hybrid
architecture for medical image classification based on neutral-atom quantum processors. 
This approach is designed to address the challenges of
medical imaging, with a particular focus on tasks such
as polyp detection and classification.
By integrating an autoencoder guided by a quantum 
reservoir, the pipeline learns compact and discriminative 
representations of image data that are also well-suited for quantum reservoir 
computing. To overcome the non-differentiability of quantum 
measurements, we circumvent this `gradient barrier' 
by incorporating a differentiable surrogate model that simulates the 
behaviour of the quantum layer, enabling end-to-end backpropagation. 
The guided training process jointly optimizes for both image 
reconstruction and classification accuracy, ensuring that the latent 
representations are both meaningful and effective for quantum processing. 
In our implementation, image data is encoded as atom detuning parameters 
in a Rydberg Hamiltonian, and quantum embeddings are obtained through expectation 
values. These embeddings are then passed to a linear classifier, 
enabling faster training and inference compared to deep classical networks.
Our experiments show that this method outperforms traditional 
approaches using PCA or unguided autoencoders. We also conduct 
ablation studies to evaluate the impact of quantum and training 
parameters, demonstrating the robustness and flexibility of the 
proposed pipeline for real-world medical imaging applications, 
even in the NISQ era.
\end{abstract}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{System Architecture Overview}
Our proposed pipeline is a Quantum-Guided Autoencoder with a 
Reservoir Surrogate (QGARS). It is a hybrid architecture that 
synergizes classical autoencoding techniques with quantum 
reservoir processing to achieve efficient learning of features
that work well with the reservoir computing layer whose embeddings
will be used to perferm a classification task.
The system is designed to overcome the inherent non-differentiability of quantum operations, which traditionally impede gradient-based training methods.
The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

In a higher level, the \textbf{architectural components} are the following:
\begin{itemize}
    \item \textbf{Classical Autoencoder:}
        \begin{itemize}
            \item \textbf{Encoder:}
            The encoder network \(\varepsilon_\omega(x)\)
            transforms high-dimensional input data \(\mathbf{x} \in \mathbb{R}^{n}\) into 
            a lower-dimensional latent representation \(\mathbf{z} \in \mathbb{R}^m\), 
            where \(m < n\).
            \item \textbf{Decoder:}
            The decoder network \(\mathcal{D}_\rho(z)\)
            Reconstructs the input data from the latent representation, producing
            \(\mathbf{\hat{x}} \in \mathbb{R}\).
        \end{itemize}

    \item \textbf{Quantum Reservoir Layer:}
        \begin{itemize}
            \item \textbf{Parameter Mapping:}
            \(g(\mathbf{z})\) maps the latent representation \(\mathbf{z}\) to the local detuning frequencies
            \(\Delta_{\mathrm{i}}\) of the Rydberg Atoms.
            \item \textbf{Quantum Evolution:}
            Evolves the system under a Hamiltonian \(H(\Delta)\), resulting in a quantum 
            state \(|\psi(\Delta)\rangle\).
            \item \textbf{Measurement:}
            Performs measurements on the quantum state to obtain embeddings \(\mathbf{q} \in \mathbb{R}^k\), which serve as training targets for the Surrogate Model or as 
            inputs for the classification task later on.
        \end{itemize}

    \item \textbf{Surrogate Model:}
    A differentiable classical neural network \( f_{\text{sur}}(\mathbf{z}; \theta_{\text{sur}}) \)
    where \(\theta_{\text{sur}}\), which are it's parameters, are tuned to approximate the 
    mapping from \(\mathbf{z}\) to \(\mathbf{q}\), effectively emulating
    the quantum layer's behaviour.

@@

\begin{figure*}[!ht]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{images/diagrams/qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{Overview of the Quantum-Guided Autoencoder Reservoir Computing System.}
\end{figure*}

@@

Ideally, the latent space representation created by the encoder \(\varepsilon_\omega(x)\)
would be passed through two sections of the model:
\begin{itemize}
    \item \textbf{The Decoder Network \(\mathcal{D}_\rho(z)\):}
    that tries to reconstruct the original input data.
    \item \textbf{The Quantum Reservoir:}
    that extracts highly dimensional quantum embeddings and
    passes them to a linear classifier to perform a classification
    task and map the embeddings to the image labels.
\end{itemize}
However, some problems emerge when using a real quantum reservoir directly during the autoencoder training stage, which will be discussed in a later section.

@@

\subsection{Rydberg Hamiltonian and Quantum Dynamics}
The quantum reservoir in our architecture is implemented
using a neutral-atom system, specifically utilizing
Rydberg atoms to create a programmable quantum reservoir.
The dynamics of the system are governed by the Rydberg Hamiltonian,
which describes the interactions between atoms in Rydberg states~\ref{eq:rydberg_hamiltonian}.

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.32\textwidth]{pca_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{ae_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{qgars_qrc_tsne.pdf}
  \caption{t-SNE projections of the 4 700-D QRC embeddings for (left) PCA + QRC, (center) AE + QRC, (right) QGARS + QRC.}
  \label{fig:embeddings_tsne}
\end{figure*}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{paper/images/results/generated_polyp_dataset/guided_lambda_log_effect.pdf}
  \caption{Accuracy as a function of the reconstruction/classification trade-off parameter $\lambda$.}
  \label{fig:lambda_sweep}
\end{figure}

@@

\begin{table}[ht]
  \caption{Test set classification accuracy (\%) for classical baselines}
  \label{tab:classical_comp}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method            & Synthetic Polyp & CVC-ClinicDB & MNIST (0/1) \\
    \hline
    PCA + Linear      &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + Linear       &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + NN           &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    \hline
  \end{tabular}
\end{table}

@@

>>>>

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{physics}
\usepackage{float}
\usepackage{inputenc}
\usepackage{bm}
%\usepackage{amsmath}

@@

% % \and
% % \IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
% % \IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
% % \textit{name of organization (of Aff.)}\\
% % City, Country \\
% % email address or ORCID}
% }

@@

\begin{abstract}
We introduce a quantum-classical hybrid
architecture for medical image classification based on neutral-atom quantum processors. This approach is designed to address the challenges of medical imaging, with a particular focus on tasks such as polyp detection and classification. By integrating an autoencoder guided by a quantum reservoir, the pipeline learns compact and discriminative representations of image data that are also well-suited for quantum reservoir computing. To overcome the non-differentiability of quantum measurements, we circumvent this `gradient barrier' by incorporating a differentiable surrogate model that simulates the behaviour of the quantum layer, enabling end-to-end backpropagation. The guided training process jointly optimizes for both image reconstruction and classification accuracy, ensuring that the latent representations are both meaningful and effective for quantum processing. In our implementation, image data is encoded as atom detuning parameters in a Rydberg Hamiltonian, and quantum embeddings are obtained through expectation values. These embeddings are then passed to a linear classifier, enabling faster training and inference compared to deep classical networks. Our experiments show that this method outperforms traditional approaches using PCA or unguided autoencoders. We also conduct ablation studies to evaluate the impact of quantum and training parameters, demonstrating the robustness and flexibility of the proposed pipeline for real-world medical imaging applications, even in the NISQ era.

\end{abstract}

@@

%============================================
% METHODOLOGY
%============================================
\section{Methodology}
\subsection{Proposed System Architecture}
Our proposed pipeline is a Quantum-Guided Autoencoder with a 
Reservoir Surrogate (QGARS). It is a hybrid architecture that 
synergizes classical autoencoding techniques with quantum reservoir processing to achieve efficient learning of features. These are expected to better match with the reservoir computing layer, whose embeddings are used to perform a classification task.

The system is designed to overcome the inherent non-differentiability of quantum operations, which traditionally impede gradient-based training methods.
The overall architecture is illustrated in `Fig.-\ref{fig:qgars_pipeline}'.

In a higher level, the \textbf{architectural components} are the following:
\begin{itemize}
    \item \textbf{Classical Autoencoder:}
        \begin{itemize}
            \item \textbf{Encoder:}
            The encoder network \(\varepsilon_\omega(\mathbf{x})\)
            transforms high-dimensional input data \(\mathbf{x} \in \mathbb{R}^{n}\) into 
            a lower-dimensional latent representation \(\mathbf{z} \in \mathbb{R}^m\), 
            where \(m < n\).
            \item \textbf{Decoder:}
            The decoder network \(\mathcal{D}_\rho(\mathbf{z})\)
            reconstructs the input data from the latent representation, producing
            \(\mathbf{\hat{x}} \in \mathbb{R}^{n}\).
        \end{itemize}

    \item \textbf{Quantum Reservoir Layer:}
        \begin{itemize}
            \item \textbf{Parameter Mapping:}
            \(g(\mathbf{z})\) maps the latent representation \(\mathbf{z}\) to the local detuning frequencies
            \(\Delta_{\mathrm{i}}\) of the Rydberg Atoms.
            \item \textbf{Quantum Evolution:}
            Evolves the system under a Hamiltonian \(H(\mathbf{\Delta)}\), resulting in a quantum 
            state \(|\psi(\mathbf{\Delta})\rangle\).
            \item \textbf{Measurement:}
            Performs measurements on the quantum state to obtain embeddings \(\mathbf{q} \in \mathbb{R}^k\), which serve as training targets for the Surrogate Model or as 
            inputs for the classification task later on.
        \end{itemize}

    \item \textbf{Surrogate Model:}
    A differentiable classical neural network \( f_{\text{sur}}(\mathbf{z}; \bm{\theta}_{\text{sur}}) \),
    where \(\bm{\theta}_{\text{sur}}\) are parameters tuned to approximate the 
    mapping from \(\mathbf{z}\) to \(\mathbf{q}\), effectively emulating
    the quantum layer's behaviour.

@@

\begin{figure*}[!t]
    \centering
    \resizebox{0.9\textwidth}{!}{
        \includegraphics{images/diagrams/qgars_pipeline_cropped.pdf}
    }
    \label{fig:qgars_pipeline}
    \caption{
    Overview of the QGARS pipeline. (a) Classical autoencoder: $\varepsilon_{\omega}(\mathbf{x})$ is the encoder network that maps input $\mathbf{x}$ to latent code $\mathbf{z}$, and $\mathcal{D}_{\rho}(\mathbf{z})$ is the decoder network reconstructing $\hat{\mathbf{x}}$. (b) Rydberg atom quantum reservoir: the latent code $\mathbf{z}$ is mapped to detuning parameters $\Delta_i$, the system evolves under the Hamiltonian in Eq.~\eqref{eq:rydberg_hamiltonian}, and measurements of observables $\langle Z_i\rangle,\langle Z_iZ_j\rangle,\langle Z_iZ_jZ_k\rangle$ yield quantum embeddings. (c) Surrogate model: a feedforward neural network $f_{\mathrm{sur}}(\mathbf{z};\theta_{\mathrm{sur}})$ trained every $N$ epochs to approximate the quantum embeddings. (d) Linear classifier: maps surrogate outputs $\mathbf{q}'$ to class predictions, closing the loop with total loss $\mathcal{L}=(1-\lambda)\mathcal{L}_R+\lambda\mathcal{L}_C$.
    }
\end{figure*}

@@

Ideally, the latent space representation created by the encoder \(\varepsilon_\omega(x)\)
would be passed through two sections of the model:
\begin{itemize}
    \item \textbf{The Decoder Network \(\mathcal{D}_\rho(z)\),}
    which tries to reconstruct the original input data.
    \item \textbf{The Quantum Reservoir,}
    which extracts highly dimensional quantum embeddings and
    passes them to a linear classifier to perform a classification
    task and map the embeddings to the image labels.
\end{itemize}
However, some problems emerge when using a real quantum reservoir directly during the autoencoder training stage, which will be discussed in a later section.

@@

\subsection{Rydberg Hamiltonian and Quantum Dynamics}
Although it can seem that the surrogate model is enough to cover the job of the quantum reservoir, that is far from the truth. while a fixed classical reservoir can nonlinearly mix inputs, its capacity is polynomially bounded. In contrast, our neutral-atom reservoir leverages many-body quantum dynamics to encode data in a space whose dimension grows exponentially with atom number, unlocking patterns that classical reservoirs would need unreasonably complex architectures to approximate. This means that the quantum reservoir is a crucial part of our pipeline and that the surrogate is just an approximation to aid in the autoencoder training process.

The quantum reservoir in our architecture is implemented
using a neutral-atom system, specifically utilizing
Rydberg atoms to create a programmable quantum reservoir.
The dynamics of the system are governed by the Rydberg Hamiltonian,
which describes the interactions between atoms in Rydberg states~\ref{eq:rydberg_hamiltonian}.

@@

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.32\textwidth]{pca_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{ae_qrc_tsne.pdf}
  \includegraphics[width=0.32\textwidth]{qgars_qrc_tsne.pdf}
  \caption{3-dimensional t-SNE projections of the 4768-dimensional QRC embeddings for (left) PCA + QRC, (center) AE + QRC, (right) QGARS + QRC}
  \label{fig:embeddings_tsne}
\end{figure*}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[width=1\linewidth]{paper/images/results/generated_polyp_dataset/guided_lambda_log_mu_effect.pdf}
  \caption{Accuracy as a function of the reconstruction/classification trade-off parameter $\lambda$.}
  \label{fig:lambda_sweep}
\end{figure}

@@

\begin{table}[ht]
  \caption{Test set classification accuracy (\%) for classical baselines}
  \label{tab:classical_comp}
  \centering
  \begin{tabular}{lccc}
    \hline
    Method            & Synthetic Polyp & CVC-ClinicDB & MNIST (0/1) \\
    \hline
    PCA + Linear      &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + Linear       &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    PCA + NN          &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    AE + NN           &  0.0 ± 0.0     &  0.0 ± 0.0   &  0.0 ± 0.0    \\
    \hline
  \end{tabular}
\end{table}

@@

% ADICIONAR COMENTÁRIO SOBRE COMO É MELHOR ESTUDAR MAIS O PARÂMETRO DE LAMBDA
